### 2022-02-08 paper/output/document/thesis_2_methods.tex
all i do is clean methods

%
\begin{isabellebody}%
\setisabellecontext{thesis{\isacharunderscore}{\isadigit{2}}{\isacharunderscore}methods}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsection{System Components \label{methods}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
My system consists of three major components: an ethical theory (Kantian ethics), a logic in which
I formalize this ethical theory (Dyadic Deontic Logic), and an interactive theorem prover in which I 
implement the formalized ethical theory (Isabelle/HOL). In this section, I describe these components, 
present the philosophical, logic, and computational background that undergirds my system, and explain
the consequences of each of the three choices I make.

These specific components determine the features
and limitations of the specific implementation of automated ethics that I present, but other choices of 
components, such as another ethical theory, a different logic, or a different theorem prover could be 
made. Given that my system is a proof-of-concept, I choose system components that are natural and 
intuitively fit together, but other choice could also be valid and perhaps even superior.

I do not claim
that I have chosen the best ethical theory or the best logic or the best interactive theorem prover, 
and flaws with these components merely demonstrate that my specific choices were incorrect, but do not 
indict logic-programming-based automated ethics as I implement it in this thesis. My thesis seeks to 
both present a specific implementation of automated ethics but also to argue for a particular approach 
to automating ethical reasoning and these choices are relevant to the former goal but not to the latter.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{Kantian Ethics \label{whykant}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In this thesis, I automate Kantian ethics. In 2006, Powers posited that deontological theories are 
attractive candidates for automation because rules are generally computationally tractable \cite[1]{powers}. 
Intuitively, algorithms are rules or procedures for problem solving and deontology offers one such 
procedure for the problem of making ethical judgements. I will make this intuition precise by
arguing that deontological ethics is natural to formalize because rules generally require little additional
data about the world and are usually easy to represent to a computer. All ethical traditions have debates that an 
automated ethical system will need to take a stance on, but these debates are less frequent and controversial
for deontological ethics than for consequentialism and virtue ethics.

I do not aim to show that deontology is the only tractable theory to automate or
to present a comprehensive overview of all consequentialist or virtue ethical theories. Instead, I 
present a sample of some approaches in each tradition and argue that deontology is more straightforward 
to formalize than these approaches. Future work could and should address the challenges I outline in 
this section. The more ethical theories that computational tools can handle, the more valuable 
computational philosophy becomes both for philosophers and for AI agents. Insofar as my project serves 
as an early proof-of-concept for computational ethics, I choose to automate an ethical theory that 
poses fewer challenges than others.

I first present deontological ethics, then consequentialism, and finally virtue ethics. For each 
tradition, I present a crash course for non-philosophers and then explain some obstacles to automation, 
arguing that these obstacles are weakest in the case of deontology. Finally, I will present the specific 
deontological theory I am automating (Kantian ethics) and will argue that it is comparatively easier
to formalize. I will also outline the specific debates in the literature that my formalization takes
a stance on and potential challenges for formalizing deontology.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Deontological Ethics\label{deontology}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
Deontological ethical theories evaluate actions as permissible, obligatory, or prohibited. The 
deontological tradition argues that an action should not be judged on its consequences, 
but rather on ``its confirmity with a moral norm" \citep{sepdeont}. In other words, deontological theories
define a set of moral norms or rules and evaluate actions using
these rules. Deontologists do not believe that we should maximize the number of times that we 
conform to such rules, but instead that we should never violate any of the moral laws. A wrong
choice is wrong, regardless of its consequences.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Deontology is immediately an attractive candidate for formalization because computers tend to 
understand rules; programming languages are designed to teach computers algorithms. Deontological
ethical theories give inviolable rules that an automated agent can apply. Moreover, because deontological 
theories focus on
the action itself, they require relatively little data. A deontological moral judgement does not require
as much information about context, consequences, or moral character as the other theories presented later in this
section. All that matters is the action and some limited set of circumstances in which it is performed.
I will later argue that, in the case of the specific deontological ethic that I implement (Kantian ethics), 
the action's representation is space efficient.

@@

While these debates are open, ``if any philosopher is regarded as central
to deontological moral theories, it is surely Immanuel Kant" \citep{sepdeont}. Out of the three ethical
traditions considered in this section, deontology has the most central representative in the form of 
Kant. Many modern deontologists agree on Kant's ethic but disagree in how to interpret it. In this paper, I will 
formalize Kantian ethics. Deontology's comparatively greater focus on Kant means that the choice of 
Kant as a guiding figure will be less controversial to deontologists than, for example, the choice of Bentham
as the guiding figure of consequentialism. Moreover, at the end of this section, I also argue that 
internal debates in the part of Kantian ethics that I focus on tend to be less controversial than those
in the consequentialist or virtue ethical traditions.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Consequentialism%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
A consequentialist ethical theory is an ethical theory that evaluates an 
action by evaluating its consequences.\footnote{There is long debate about what exactly makes an ethical theory consequentialist \citep{consequentialismsep}. 
For this thesis, I focus on theories that place the moral worth of an act in its the consequences.} For example, 
utilitarianism is a form of consequentialism in which the moral action 
is the action that produces the most good \citep{utilsep}. The focus
on the consequences of action distinguishes consequentialists from deontologists, who derive the moral worth
of an action from the action itself. Some debates in the consequentialist tradition include 
which consequences of an action matter, what exactly constitutes a ``good" consequence, and how we can 
aggregate the consequences of an action over all the individuals involved.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Which Consequences Matter}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Because consequentialism evaluates the state of affairs following an action, this kind of ethical 
reasoning requires more knowledge about the state of the world than deontology. Consequentialism
requires knowledge about some or all consequences following an action. This requires that an automated 
ethical system somehow collect a subset of the infinite consquences of following an action, a difficult, 
if not impossible, task. Moreover, compiling this database of consequences requires 
answering difficult questions about which consequences were actually caused by an action.\footnote
{David Hume argues that many straightforward accounts of causation face difficulties \citep{hume}, 
and philosohpers continue to debate the possiblity of knowing an event's true cause. Kant even argued
that first causes, or noumena, are unknowable by human beings \citep{kantnoumena}.}
Evaluating an effect of an action requires knowledge about the 
state of the world before and after an action and knowledge about the action itself. Consequentialism requires 
knowledge about the situation in which the act is performed and following the act, whereas deontology mostly requires 
knowledge about the act itself. As acts become
more complex and affect more people, the computational time and space required to calculate and store
their consequences increases. Deontology, on the other hand, does not suffer this scaling
challenge because acts that affect 1 person and acts that affect 1 million people share the same
representation.

The challenge of representing the circumstances of action is not unique to consequentialism, but is particularly acute in this case. Kantian ethicists robustly debate which circumstances of an action are ``morally relevant'' when evaluating an action's moral worth.\footnote{Powers 
\citet{powers} identifies this as a challenge for automating Kantian ethics and briefly sketches 
solutions from \citet{constofreason}, \citet{silber}, and \citet{rawlsconstructivism}. For more on 
morally relevant circumstances, see Section WhatIsAMaxim.} Because deontology merely evaluates a 
single action, the surface of this debate is much smaller than the debate about circumstances and 
consequences in a consequentialist system. An automated consequentialist system must make such 
judgements about the act itself, the circumstances in which it is performed, and the circumstances 
following the act. All ethical theories relativize their judgements to the situation in which an act 
is performed, but consequentialism requires far more knowledge about the world than deontology.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Theory of the Good}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
An automated consequentialist reasoner must also take a stance on the debate over
what qualifies as a ``good consequence," or what the theory of the good is. For example, hedonists associate
good with the presence of pleasure and the absence of pain, while preference utiliarians believe that good is 
the satisfaction of individuals' desires. Other consequentialists, like Moore, adopt a pluralistic theory of value, under which 
many different kinds of things are good for different reasons \citep{moorepe}.

Most of the above theories of good require that a moral reasoner understand complex features about
individuals' preferences, desires, or sensations in order to evaluate a moral action, making automated
consequentialist ethics difficult. Evaluating a state of affairs requires
judgements about whether a state of affairs actually satisifes the relevant criteria for goodness. 
These judgements are controversian, and any consequentialist decision requires many of 
these judgements for each individual involved. As systems involve more people, making these judgements 
quickly becomes difficult, posing a scaling challenger. Perfect knowledge of tens of thousands of 
people's pleasure or preferences or welfare or rights is impossible. Either a human being 
assigns values to states of affairs, which doesn't scale, or the machine does, 
which requires massive common-sense and increases room for doubting the system's judgements. This is 
a tractable problem, but it is much more difficult than the equivalent deontological task of formulating
and evaluating an action.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Aggregation}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Once an automated consequentialist agent assigns a goodness measurement to each person in a state of affairs, it 
must also calculate an overall goodness measurement for the state of affairs. One approach to assigning
this value is to aggregate each person's individual goodness score into one complete score for a state. 
For example, under a simple welfare model, each person is assigned a welfare score and the total 
score for a state of affairs is the sum of the welfare scores for each involved person.
The more complex the theory of the good, the more difficult this aggregation becomes. For example, 
pluralistic theories struggle to explain how different kinds of value can be compared \citep{consequentialismsep}. 
How do we compare one unit of beauty to one unit of pleasure? Subjective theories of the good, such 
as those focused on the sensation of pleasure or an individual's preferences, present difficulties in 
comparing different people's subjective measures. Resolving this debate requires that the automated reasoner 
choose one specific aggregation algorithm, but those who disagree with this choice will not trust 
the reasoner's moral judgements. Moreover, for complex theories of the good, this aggregation algorithm
may be complex and may require a lot of data.

@@

evaluations of a state of affairs. While this approach no longer requires that a reasoner define an 
aggregation algorithm, the reasoner still needs to calculate a goodness measurement for a state of 
affairs. Whereas before the reasoner could restrict analysis to a single person, the algorithm must now 

@@

``discount factor" $\gamma < 1$, such that if $r_i$ is the reward at time $i$, the total reward is $\sum_{i=0}^{\infty}\gamma^i r_i$.} 

@@

similar problems because, at some point, a useful automated consequentialist moral agent will need 

@@

their possessor good, at the center \cite{vesep}. For example, Aristotle describes virtues as the 

@@

ethicists present a list of virtues. Such theories vary from Aristotle's virtues of courage and 
temperance \citet{aristotle} to the Buddhist virtue of equanimity \citet{mcrae}. An automated virtue 

@@

Another difficulty with automating virtue ethics is that the unit of evaluation for a virtue ethical
theory is often a person's entire moral character. While deontologists evaluate the act itself and utilitarians
evaluate the consequences of an act, virtue ethicists evaluate the actor's moral character and their 
disposition towards the act. Virtues are character traits and evaluating an action as virtuous or 
not requires understanding the agent's character and disposition while acting. If states of affairs

@@

to represent to a computer. Consequentialism posed a data-collection problem in evaluating and representing states
of affairs, but virtue ethics poses a conceptual problem about the formal nature of moral character.

@@

One potential appeal of virtue ethics is that many virtue ethical theories involve some form of 

@@

imply that ethical behavior can be learned from some dataset of virtuous acts, either those 
prescribed by a moral teacher or those that a virtuous ideal agent would undertake. Indeed, these 
theories seem to point towards a machine learning approach to computational ethics, in which ethics is 

@@

Machine learning approaches like the Delphi system \citep{delphi} mentioned in \ref{intro} also may suffer explanability 
problems that my logic-programming, theorem-prover
approach does not face. Many machine learning algorithms cannot sufficiently explain their 
decisions to a human being, and often find patterns or correlations in datasets that don't actually 
cohere with the trends and causes that a human being would identify \citep{puiutta}. While there is significant activity 
and progress in explainable machine learning, interactive theorem provers are designed to be explainable 
at the outset. Indeed, Isabelle can show the axioms and lemmas it used in constructing a proof, 
allowing a human being to reconstruct the proof independently if they wish. This is not an 
intractable problem for machine learning approaches to computational ethics, but is one reason to 
prefer logical approaches.\footnote{This argument about explanability is in the context of virtue ethics and 
machine learning. It also applies to a broader class of work in automated ethics 
that uses ``bottom-up" approaches, in which a system learns moral judgements from prior judgements. 
I will extend this argument to general bottom-up approaches in Section Related Work.}%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Kantian Ethics \label{kantianethics}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In this paper I focus on Kantian ethics, a specific branch of deontology. Kant's theory is centered 
on practical reason, which is the kind of reason that we 
use to decide what to do. In \emph{The Groundwork of the Metaphysics of Morals}, Kant explains that 
rational beings are unique because we can act ``in accordance with 
the representations of laws" \citep[4:412]{groundwork}. In contrast, a ball thrown into the air acts 
according to the laws of physics. It cannot ask itself, ``Should I fall back to the ground?" 
It simply falls. A rational being, on the other hand, can ask, ``Should I act on this reason?" 
As Korsgaard describes it, when choosing which desire to act on, ``it is as if there is something over 
and above all of your desires, something which is you, and which chooses which desire to act on" \citep[100]{sources}. 
Rational beings are set apart by this reflective capacity. A rational being's behavior is purposive and 
their actions are guided by practical reason. They have reasons for acting, even when these reasons are
opaque to them. This reflective choosing, or operation of practical reason, is what Kant calls the will.

The will operates by adopting, or willing, maxims, which are its perceived reasons for acting. Kant defines a maxim as 
the ``subjective principle of willing," or the reason that the will \emph{subjectively} gives 
to itself for acting \citep[16 footnote 1]{groundwork}. Many philosophers agree that a maxim consists 
of some combination of circumstances, 
act, and goal.\footnote{For more discussion of the definition of a maxim, see Section What Is a Maxim}
One example of a maxim is ``When I am hungry, I will eat a doughnut in order to satisfy my sweet tooth." 
When an agent wills this maxim, they decide to act on it. They commit themselves to the end in the maxim 
(e.g. satisfying your sweet tooth). They represent their action, to themselves, as following the 
principle given by this maxim. Because a maxim captures an agent's principle of action, Kant evaluates
maxims as obligatory, prohibited, or permissible. He argues that the form of certain maxims 
requires any rational agent to will them, and these maxims are obligatory.

The form of an obligatory maxim is given by the categorical imperative. 
An imperative is a command, such as ``Close the door" or ``Eat the doughnut in order to satisfy your 
sweet tooth." An imperative is categorical if it holds unconditionally for all rational agents under all 
circumstances. Kant argues that the moral law must be a categorical imperative \citep[5]{groundwork}. 
In order for an imperative to be categorical, it must be derived from the will's authority over itself. 
Our wills are autonomous, so the only thing that can have unconditional authority over a rational will is 
the will itself. No one else can tell you what to do because you can always ask why you 
should obey their authority. The only authority that you cannot question is the authority of your own 
practical reason. To question this authority is to demand a reason for acting for reasons, which 
concedes the authority of reason itself \citep[23]{velleman}. Therefore, the only possible candidates 
for the categorical imperative are those rules that are required of the will because it is a will.

Armed with this understanding of practical reason, Kant presents the categorical 
imperative. He presents three ``formulations" or versions of the categorical imperative and goes on to 
argue that all three formulations are equivalent. In this project, I focus on the first formulation,
the Formula of Universal Law, and I explain this choice in Section \ref{whyful}.

The first formulation of the categorical imperative is the
Formula of Universal Law (FUL), which reads, ``act only according to that maxim through which you can 
at the same time will that it become a universal law" \cite[34]{groundwork}. This formulation
generates the universalizability test, which tests the moral value of a maxim by 
imagining a world in which it becomes a universal law and attempting to will the maxim in that world.
If there is a contradiction in willing the maxim in a world in which everyone universally wills the maxim,
the maxim is prohibited.

Velleman presents a concise argument for the FUL. He argues that reason is universally shared among reasoners. For 
example, all reasoners have equal access to the arithmetic logic that shows that ``2+2=4" \cite[29]{velleman}. The chain of 
reasoning that makes this statement true is not specific to any person, but is universal across people. 
Therefore, if I have sufficient reason to will a maxim, so does every other rational agent. There is 
nothing special about the operation of my practical reason that other reasoners don't have access to. 
In adopting a maxim, I implicitly state that all reasoners
across time also have reason to adopt that maxim. Therefore, because I act on reasons, I must obey the 
FUL. Notice that this fulfills the above criterion for a categorical imperative: the FUL is derived from 
a property of practical reason itself and thus derives authority from the will's authority over itself, 
as opposed to some external authority.

>>>>

%
\begin{isabellebody}%
\setisabellecontext{thesis{\isacharunderscore}{\isadigit{2}}{\isacharunderscore}methods}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isatagtheory
%
\endisatagtheory
{\isafoldtheory}%
%
\isadelimtheory
%
\endisadelimtheory
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsection{System Components \label{methods}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
My system consists of three components: an ethical theory (Kantian ethics), a logic in which
I formalize this ethical theory (Dyadic Deontic Logic), and an interactive theorem prover in which I 
implement the formalized ethical theory (Isabelle/HOL). In this section, I describe these components, 
present the philosophical, logic, and computational background underlying my system, and explain
the consequences of each of the three choices I make.

These specific components determine the features
and limitations of my implementation of automated ethics, but other choices of 
components, such as another ethical theory, a different logic, or a different theorem prover could be 
made. Flaws with these components are merely limitations of my system, but do not 
indict logic-programming-based automated ethics more generally. My thesis seeks to 
both present a specific implementation of automated ethics but also to argue for a particular approach 
to automating ethical reasoning and these choices are relevant to the former goal but not to the latter.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{Kantian Ethics \label{whykant}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In this thesis, I automate Kantian ethics. In 2006, Powers posited that deontological theories are 
attractive candidates for automation because rules are generally computationally tractable \cite[1]{powers}. 
Intuitively, algorithms are rules or procedures for problem solving and deontology offers one such 
procedure for the problem of making ethical judgements. I will make this intuition precise by
arguing that deontological ethics is natural to formalize because rules generally require little additional
data about the world and are usually easy to represent to a computer. All ethical traditions have debates that an 
automated ethical system will need to take a stance on, but these debates are less frequent and controversial
for deontological ethics than for consequentialism and virtue ethics.

I do not aim to show that deontology is the only tractable theory to automate or
to present a comprehensive overview of all consequentialist or virtue ethical theories. Instead, I 
present a sample of some approaches in each tradition and argue that deontology is more straightforward 
to formalize than these approaches. Insofar as my project serves 
as an early proof-of-concept, I choose to automate an ethical theory that 
poses fewer challenges than others.

I first present deontological ethics, then consequentialism, and finally virtue ethics. For each 
tradition, I present a crash course for non-philosophers and then explain some obstacles to automation, 
arguing that these obstacles are weakest in the case of deontology. Finally, I will present the specific 
deontological theory I am automating (Kantian ethics) and will argue that it is comparatively easier
to formalize.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Deontological Ethics\label{deontology}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
Deontological theories evaluate actions as permissible, obligatory, or prohibited and judge actions 
not on their consequences, but rather on ``confirmity with a moral norm" \citep{sepdeont}. In other words, 
deontological theories define a set of moral rules and evaluate actions using
these rules. Deontologists believe that we should never violate any moral law. A wrong
choice is wrong, regardless of its consequences.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Deontology is an attractive candidate for formalization because computers tend to 
understand rules; programming languages are designed to teach computers algorithms. Deontological
ethical theories give inviolable rules that an automated agent can apply. Moreover, because deontological 
theories focus on the action itself, they require relatively little data. A deontological moral judgement does not require
as much information about context, consequences, or moral character as the other theories presented later in this
section. All that matters is the action and some limited set of circumstances in which it is performed.

@@

While these debates are open, ``if any philosopher is regarded as central
to deontological moral theories, it is surely Immanuel Kant" \citep{sepdeont}. Out of the three ethical
traditions considered in this section, deontology has the most central representative in the form of 
Kant. In this paper, I will 
formalize Kantian ethics. Deontology's comparatively greater focus on Kant means that the choice of 
Kant as a guiding figure will be less controversial to deontologists than, for example, the choice of Bentham
as the guiding figure of consequentialism. Moreover, at the end of this section, I also argue that 
internal debates in the part of Kantian ethics that I focus on tend to be less controversial than those
in the consequentialist or virtue ethical traditions.%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Consequentialism%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
A consequentialist ethical theory is an ethical theory that evaluates an 
action by evaluating its consequences.\footnote{There is long debate about what exactly makes an ethical theory consequentialist \citep{consequentialismsep}. 
For this thesis, I focus on theories that place the moral worth of an act in its the consequences.} For example, 
utilitarianism is a form of consequentialism in which the moral action 
is the action that produces the most good \citep{utilsep}. The focus
on the consequences of action distinguishes consequentialists from deontologists, who derive the moral worth
of an action from the action itself. Some debates in the consequentialist tradition include 
which consequences matter, what constitutes a ``good" consequence, and how we can 
aggregate the consequences of an action over all the individuals involved.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Which Consequences Matter}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Because consequentialism evaluates the state of affairs following an action, this kind of ethical 
reasoning requires more knowledge about the state of the world than deontology. Consequentialism
requires knowledge about some or all consequences following an action. This requires that an automated 
ethical system somehow collect a subset of the infinite consquences of following an action, a difficult, 
if not impossible, task. Moreover, compiling this database of consequences requires 
answering difficult questions about which consequences were actually caused \footnote
{David Hume argues that many straightforward accounts of causation face difficulties \citep{hume}, 
and philosophers continue to debate the possiblity of knowing an event's true cause. Kant even argued
that first causes, or noumena, are unknowable by human beings \citep{kantnoumena}.} by an action and
determining the state of the world before and after an action. As acts become
more complex and affect more people, the computational time and space required to calculate and store
their consequences increases. Deontology, on the other hand, does not suffer this scaling
challenge because acts that affect 1 person and acts that affect 1 million people share the same
representation.

The challenge of representing the circumstances of action is not unique to consequentialism, but is particularly acute in this case. 
Kantian ethicists robustly debate which circumstances of an action are ``morally relevant'' when evaluating an action's moral worth.\footnote{ 
\citet{powers} identifies this as a challenge for automating Kantian ethics and briefly sketches 
solutions from \citet{constofreason}, \citet{silber}, and \citet{rawlsconstructivism}. For more on 
morally relevant circumstances, see Section WhatIsAMaxim.} Because deontology merely evaluates a 
single action, the surface of this debate is much smaller than debates about circumstances and 
consequences in a consequentialist system. An automated consequentialist system must make such 
judgements about the act itself, the circumstances in which it is performed, and the circumstances 
following the act. All ethical theories relativize their judgements to the situation in which an act 
is performed, but consequentialism requires far more knowledge about the world than deontology.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Theory of the Good}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
An automated consequentialist reasoner must also take a stance on the debate over
what qualifies as a ``good consequence,'' or what the theory of the good is. For example, hedonists associate
good with the presence of pleasure and the absence of pain, while preference utiliarians believe that good is 
the satisfaction of desire. Other consequentialists, like Moore, adopt a pluralistic theory of value, under which 
many different kinds of things are good for different reasons \citep{moorepe}.

Most theories of the good require that a moral reasoner understand complex features about
individuals' preferences, desires, or sensations in order to evaluate a moral action, making automated
consequentialist ethics difficult. Evaluating a state of affairs requires many controversial
judgements about whether a state of affairs actually satisifes the relevant criteria for goodness. 
Perfect knowledge of tens of thousands of 
people's pleasure or preferences or welfare or rights is impossible. Either a human being 
assigns values to states of affairs, which doesn't scale, or the machine does, 
which requires massive common-sense and increases room for doubting the system's judgements. This may be 
a tractable problem, but it is much more difficult than the equivalent deontological task of formulating
and evaluating an action.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Aggregation}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Once an automated consequentialist agent assigns a goodness measurement to each person in a state of affairs, it 
must also calculate an overall goodness measurement for the state of affairs. One approach to assigning
this value is to aggregate each person's individual goodness score into one complete score for a state. 
The more complex the theory of the good, the more difficult this aggregation becomes. For example, 
pluralistic theories struggle to explain how different kinds of value can be compared \citep{consequentialismsep}. 
How do we compare one unit of beauty to one unit of pleasure? Resolving this debate requires that an automated reasoner 
choose one specific aggregation algorithm, but those who disagree with this choice will not trust 
the reasoner's moral judgements. Moreover, for complex theories of the good, this aggregation algorithm
may be complex and may require a lot of data.

@@

evaluations of a state of affairs. While this approach no longer requires that an 
aggregation algorithm, an automated ethical system still needs to calculate a goodness measurement for a state of 
affairs. Whereas before the system could restrict its analysis to a single person, the algorithm must now 

@@

``discount factor'' $\gamma < 1$, such that if $r_i$ is the reward at time $i$, the total reward is $\sum_{i=0}^{\infty}\gamma^i r_i$.} 

@@

similar problems because, at some point, a usable automated consequentialist moral agent will need 

@@

their possessor good, at the center \citep{vesep}. For example, Aristotle describes virtues as the 

@@

ethicists present a list of virtues, such as the Buddhist virtue of equanimity \citep{mcrae}. An automated virtue 

@@

Another difficulty with automating virtue ethics is that the unit of evaluation for virtue ethics
is often a person's entire moral character. While deontologists evaluate the act itself, virtue ethicists 
evaluate the actor's moral character and their 
disposition towards the act. If states of affairs

@@

to represent to a computer. This is more than just a data-collecting problem; it is a conceptual problem 
about the formal nature of moral character.

@@

One potential appeal of virtue ethics is that many virtue ethical theories involve some notion of 

@@

implies that ethical behavior can be learned from some dataset of virtuous acts, either those 
prescribed by a moral teacher or those that a virtuous ideal agent would undertake. These 
theories seem to point towards a machine learning approach to automated ethics, in which ethics is 

@@

Machine learning approaches like the Delphi system \citep{delphi} mentioned in Chapter \ref{intro} also may suffer explanability 
problems that my logic-programming, theorem-prover
approach does not face. Many machine learning algorithms cannot sufficiently explain their 
decisions to a human being, and often find patterns in datasets that don't 
cohere with the causes that a human being would identify \citep{puiutta}. While there is significant activity 
and progress in explainable machine learning, interactive theorem provers are designed to be explainable 
at the outset. Isabelle can show the axioms and lemmas it used in constructing a proof, 
allowing a human being to reconstruct the proof independently if they wish. This is not an 
intractable problem for machine learning approaches to computational ethics, but is one reason to 
prefer logical approaches.\footnote{This argument about explanability is in the context of virtue ethics and 
machine learning. It also applies to a broader class of work in automated ethics 
that uses ``bottom-up" approaches, in which a system learns moral judgements from prior judgements. 
I will extend this argument to general bottom-up approaches in Section Related Work.}%
\end{isamarkuptext}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsubsection{Kantian Ethics \label{kantianethics}%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
In this paper I focus on Kantian ethics, a specific branch of deontology. Kant's theory is centered 
on practical reason, which is the kind of reason that we 
use to decide what to do and the source of our agency. In \emph{The Groundwork of the Metaphysics of Morals}, Kant explains that 
rational beings are unique because we act ``in accordance with 
the representations of laws'' \citep[4:412]{groundwork}. In contrast, a ball thrown into the air acts 
according to the laws of physics. It cannot ask itself, ``Should I fall back to the ground?'' 
It simply falls. A rational being, on the other hand, can ask, ``Should I act on this reason?"''
As Korsgaard describes it, when choosing which desire to act on, ``it is as if there is something over 
and above all of your desires, something which is you, and which chooses which desire to act on'' \citep[100]{sources}. 
Rational beings are set apart by this reflective capacity. We are purposive and 
our actions are guided by practical reason. We have reasons for acting, even when these reasons are
opaque to us. This reflective choosing, or operation of practical reason, is what Kant calls the will.

The will operates by adopting, or willing, maxims, which are its perceived reasons for acting. Kant defines a maxim as 
the ``subjective principle of willing,'' or the reason that the will \emph{subjectively} gives 
to itself for acting \citep[16 footnote 1]{groundwork}. Many philosophers agree that a maxim consists 
of some combination of circumstances, 
act, and goal.\footnote{For more discussion of the definition of a maxim, see Section What Is a Maxim}
One example of a maxim is ``When I am hungry, I will eat a doughnut in order to satisfy my sweet tooth.''
When an agent wills this maxim, they decide to act on it. They commit themselves to the end in the maxim 
(e.g. satisfying your sweet tooth). They represent their action, to themselves, as following the 
principle given by this maxim. Because a maxim captures an agent's principle of action, Kant evaluates
maxims as obligatory, prohibited, or permissible. He argues that the form of certain maxims 
requires any rational agent to will them, and these maxims are obligatory.

The form of an obligatory maxim is given by the categorical imperative. 
An imperative is a command, such as ``Close the door" or ``Eat the doughnut in order to satisfy your 
sweet tooth." An imperative is categorical if it holds unconditionally for all rational agents in all 
circumstances. Kant argues that the moral law must be a categorical imperative \citep[5]{groundwork}. 
In order for an imperative to be categorical, it must be derived from the will's authority over itself. 
Our wills are autonomous, so the only thing that can have unconditional authority over a rational will is 
the will itself. No one else can tell you what to do because you can always ask why you 
should obey their authority. The only authority that you cannot question is the authority of your own 
practical reason. To question this authority is to demand a reason for acting for reasons, which 
concedes the authority of reason itself \citep[23]{velleman}. Therefore, the only possible candidates 
for the categorical imperative are those rules that are required of the will because it is a will.

Armed with this understanding of practical reason, Kant presents the categorical 
imperative. He presents three ``formulations" or versions of the categorical imperative. In this project, 
I focus on the first formulation, the Formula of Universal Law, and I explain this choice in Section \ref{whyful}.

The Formula of Universal Law (FUL) states, ``act only according to that maxim through which you can 
at the same time will that it become a universal law'' \cite[34]{groundwork}. This formulation
generates the universalizability test, which we can use to test the moral worth of a maxim by 
imagining a world in which it becomes a universal law and attempting to will the maxim in that world.
If there is a contradiction in willing the maxim in a world in which everyone universally wills the maxim,
the maxim is prohibited.

Velleman presents a concise argument for the FUL. He argues that reason is universally shared among reasoners. For 
example, all reasoners have equal access to the arithmetic logic that shows that ``2+2=4'' \cite[29]{velleman}. The 
reasoning that makes this statement true is not specific to any person, but is universal across people. 
Therefore, if I have sufficient reason to will a maxim, so does every other rational agent. There is 
nothing special about the operation of my practical reason. 
In adopting a maxim, I implicitly state that all reasoners
across time also have reason to adopt that maxim. Therefore, because I act on reasons, I must obey the 
FUL. Notice that this fulfills the above criterion for a categorical imperative: the FUL is derived from 
a property of practical reason itself and thus derives authority from the will's authority over itself.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
\noindent \textbf{Ease of Automation}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Kantian ethics is an especially candidate for formalization because the categorical imperative, particularly the FUL, 
is a property of reason related to the form or structure of a maxim. It does not require any situational 
knowledge beyond the circumstances included
in the maxim itself and thus requires fewer contingent facts than other ethical theories.
While other ethical theories often rely on many facts about 
the world or the actor, a computer evaluating 
a maxim doesn't require any knowledge about the world beyond what is contained in a maxim. Automating 
Kantian ethics merely requires making the notion of a maxim precise and representing it to the computer. 
This distinguishes Kantian ethics from consequentialism and virtue ethics, which
require far more knowledge about the world or the agent to reach a moral decision.
